Advanced Normalization Tools (ANTS)

### 2009-09-30 Documentation/Tex/ants.tex
large def
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@158 fd2da0a4-d968-48df-9cb5-d8c496287cbf

\section{Introduction}
The ANTS framework provides open-source, encapsulated,
software-engineered functionality for deformable normalization and
segmentation.  Independent evaluation of ANTS normalization tools, 
applied to ``control'' data, 
placed the toolkit as a top performer amongst 14
methods \cite{Klein2009}.  Developer evaluation showed stronger differences 
with other methodology in neurodegenerative neuroimaging data, 
where large deformation is required \cite{Avants2008b}. 
ANTS serves as both a base for further
algorithm development and also as an application-oriented toolkit.
ANTS enables diffeomorphic normalization with a variety of
transformation models, optimal template construction, multiple types
of diffeomorphisms, multivariate similarity metrics, diffusion tensor
processing and warping, image segmentation with and without priors and
measurement of cortical thickness from probabilistic segmentations.
The normalization tools, alone, provide a near limitless range of
functionality and allow the user to develop customized objective
functions.  Objective functions in ANTS are of the form: $$
\text{Deformation Cost} + \text{Data Terms},$$ and the command
line reflects this balance of two terms.  As mentioned above, the
data term may combine multiple different measures of
similarity that are optimized in parallel, for instance, image similarity 
and landmark terms.  This document seeks to
provide a practical overview of basic functionality and some of the common use cases that
users seek.
Additional information is available online -- see 
\href{www.picsl.upenn.edu/ANTS}{ANTS Homepage : www.picsl.upenn.edu/ANTS}.
For compilation details, see: \href{http://picsl.upenn.edu/ANTS/download.php}{Compile and download: http://picsl.upenn.edu/ANTS/download.php}.
The most important ANTS programs are described below. 
\footnote{This document is a work in progress. Please check for updates with each release.}

@@

\subsection{ANTS Scripts} 
The ANTS/Scripts directory contains (hopefully) user-friendly
wrappings of ANTS tools that enable higher-level error checking and
combinations of basic ANTS functions.  These scripts are called as
\verb sh ~ \verb antsscriptname.sh ~ and provide usage when called
from the command line.  For instance, try \verb sh ~ \verb ants.sh .
Additionally, if you run ants in tcsh shells, you will need to call it
as follows:
\begin{verbatim}
~/Code/ANTSSourceForgeSVN/bin/ANTS 2 -m PR\[r16slice.nii,r64slice.nii,1,2\]
\end{verbatim}
instead of
\begin{verbatim}
~/Code/ANTSSourceForgeSVN/bin/ANTS 2 -m PR[r16slice.nii,r64slice.nii,1,2]
\end{verbatim}
because tcsh does no interpret the brackets as intended.  Using the
bash shell script interface also avoids this.  The key change involves using the slash to force tcsh to 
interpet the bracket literally.

\section{ImageRegistration with ANTS}
\begin{wrapfigure}{r}{0.5\textwidth}
\framebox[0.5\textwidth]{
\begin{minipage}[t]{0.45\textwidth}
{\bf Quick Start:} {\color{blue}call ants.sh (from ANTS/Scripts/) to get usage and apply a normalization to some of your existing data.  It is instructional to read the script, modify some of the parameters and re-run to witness the effect of your changes -- image registration is an art as well as science.  Many other ``ready to go'' scripts are available in ANTS/Scripts. The user must set his/her ANTSPATH environment variable  -- which points to the location of ANTS binaries -- within these scripts or in their environment.  Note: All ANTS programs provide usage when called from the command line.  Most require the image dimension to be specified as the first parameter.  E.g. 
  ImageMath ImageDimension 
where ImageDimenion is 2 or 3.}
\end{minipage}}
\end{wrapfigure}
There are two general applications of image registration. The first application is transforming labeled data from a template image space into an individual space. This strategy is important when appearance alone is not enough to locate a structure as, for example, in the case of hippocampus segmentation. The template provides a prediction of the hippocampus-amygdala boundary.
The second application operates in the "inverse" direction of the first: instead of mapping template to individual, we map individual(s) to the template. Voxel-based population studies of either functional or structural variables depend on mapping to a template space. The common coordinate system enables a statistical evaluation of the likelihood of consistent activation across a group or, in other contexts, the differences in anatomy between two groups.

@@

\subsection{Optimal Template Construction with ANTS}
A useful script for setting up optimal template construction
 is available in ANTS/Scripts.  This method is described in \cite{Avants2004,Avants2006d,Kim2008,Yushkevich2009}, 
where more recent publications are more descriptive and incorporate 
more recent features and optimization strategies. 
Two versions of this algorithm are available : serial and parallel.  
Data and a script for building an average face template is available 
the ANTS/Examples/Data directory.  Two key points about optimal templates:
1. The outcome stabilizes at around ten images, for most populations.
That is, if we randomly select images from individuals within the same 
demographic distribution, we will end up with very similar average images.  
2.  Optimality, for the ANTS-SyN approach, is defined by the minimum shape 
and appearance distance image. 
See the template page for some examples of this and users may 
contact ANTS developers for previously derived template 
images that may be useful.

@@

@@

CreateWarpedGridImage 2 PSEtestInverseWarp.nii grid1.nii
CreateWarpedGridImage 2 PSEtestWarp.nii grid2.nii
\end{verbatim}
This example should run on the downloaded ANTS data so you may see the results.

%\input{ANTSLabeledDataExample}
%\input{ANTSMaskedImageRegistrationExample}

\section{Application to Studies}
\subsection{Brain Mapping in the Presence of Lesions}
Many cases violate the basic assumption of a one-to-one mapping from a
template image to a target image. Brain lesions caused by stroke or
traumatic brain injury are a common instance of this.

@@

\subsection{Statistics with ANTS and {\bf R}}
{\bf R} is an open-source, cutting-edge statistical package 
used by statisticians world-wide.   ANTS is designed to 
interface with {\bf R} via an ImageMath tool that converts 
image sets into an {\bf R} compatible format.  
Three steps are required.  First, convert image data into 
vector and matrix formats through a mask of your statistical 
region of interest.  Second, read that data into {\bf R} and 
run your statistics.  Third, write out the data from {\bf R} and 
convert back to an image.  Here are the steps, in code examples,
\begin{enumerate}
\item Create a vector for your output: 
\begin{verbatim} 
ImageMath 3 testresultvec.nii ConvertImageSetToMatrix 1 mask.nii image1.nii
\end{verbatim}
Here, we assume all images are called image1.nii, image2.nii, ... , imageN.nii so we 
can use a wildcard to pass to ImageMath and 
create a matrix for all your input data: 
\begin{verbatim} 
 ImageMath 3 datamatrix.nii ConvertImageSetToMatrix 1 mask.nii image*.nii 
\end{verbatim}
\item   Run {\bf R} by reading your data above, performing whatever statistics you want and then 
writing out.  
\item   Convert the test result back to an image.  
\begin{verbatim} 
ImageMath 3 testresult.nii ConvertVectorToImage mask.nii testresultvec.nii
\end{verbatim}
\end{enumerate}
Most statistical requirements may be met with this setup. 
\section{Dependencies and Related Software}
\subsection{Dependencies and Compilation}
ANTS depends on the most recent version of the Insight ToolKit (ITK)
-- www.itk.org.   File types that may be read and written are 
all of those available in itk and a few more \href{http://picsl.upenn.edu/ANTS/ioants.php}{http://picsl.upenn.edu/ANTS/ioants.php}.

@@

\subsection{Pipelining with ANTS}
Use PipeDream for automating large-scale processing:\newline
\href{https://sourceforge.net/projects/neuropipedream/}{PipeDream
  Homepage :
  https://sourceforge.net/projects/neuropipedream/} \newline PipeDream
is integrated with ANTS and automates more complex studies such as
multivariate diffusion tensor and T1 cortical thickness studies,
longitudinal mapping and reconstruction of nifti images from Dicom.

\section{Annotated Bibliography}
The original statement of the symmetric normalization and template
construction methodology was given in \cite{Avants2004}.  A follow up
study that used landmark guidance to compare the chimpanzee cortex to
the human cortex was published here \cite{Avants2006} -- this study
used {\em in vivo} MRI and template-based normalization to confirm
volumetric numbers derived from an early 20th century post-mortem
study comparing one human and one chimp.  This conference article has
some additional detail and alternative updates to the methodology, in
particular application to shape-based interpolation
\cite{Avants2005b}.  Network based studies were performed here
\cite{duda08miccai,duda08cvpr}.  The main SyN paper is here
\cite{Avants2008}.  Applications to neurodegeration are here
\cite{Avants2005,Avants2008a,Grossman2008,Avants2009,Das2009,Yushkevich2009,Massimo2009}.
Hippocampus focused work is here \cite{Pluta2009,Yushkevich2009}.  The
main evaluation papers include \cite{Avants2008} and \cite{Klein2009}
for the cortex and deep brain structures whereas \cite{Pluta2009}
evaluates the use of automated and semi-automated normalization for
high-throughput hippocampus morphometry.  An additional evaluation
paper is being developed.

>>>>

\section{Introduction}
The ANTS framework provides open-source functionality for deformable
normalization with large deformations, as shown in
figure~\ref{fig:chalf}.  Small deformation mappings and segmentation 
tools are also available. 
Independent evaluation of ANTS normalization tools, 
applied to ``control'' data, 
placed the toolkit as a top performer amongst 14
methods \cite{Klein2009}.  Developer evaluation showed stronger differences 
with other methodology in neurodegenerative neuroimaging data, 
where large deformation is required \cite{Avants2008b}. 
ANTS serves as both a base for further
algorithm development and also as an application-oriented toolkit.
ANTS enables diffeomorphic normalization with a variety of
transformation models, optimal template construction, multiple types
of diffeomorphisms, multivariate similarity metrics, diffusion tensor
processing and warping, image segmentation with and without priors and
measurement of cortical thickness from probabilistic segmentations.
The normalization tools, alone, provide a near limitless range of
functionality and allow the user to develop customized objective
functions.  Objective functions in ANTS are of the form: $$
\text{Deformation Cost} + \text{Data Terms},$$ and the command
line reflects this balance of two terms.  As mentioned above, the
data term may combine multiple different measures of
similarity that are optimized in parallel, for instance, image similarity 
and landmark terms.  This document seeks to
provide a practical overview of basic functionality and some of the common use cases that
users seek.
Additional information is available online -- see 
\href{www.picsl.upenn.edu/ANTS}{ANTS Homepage : www.picsl.upenn.edu/ANTS}.
For compilation details, see: \href{http://picsl.upenn.edu/ANTS/download.php}{Compile and download: http://picsl.upenn.edu/ANTS/download.php}.
The most important ANTS programs are described below. 
\footnote{This document is a work in progress. Please check for updates with each release.}
\begin{figure}
\includegraphics[width=0.33\textwidth]{Figures/grid1100.pdf} 
\includegraphics[width=0.33\textwidth]{Figures/grid1110.pdf} 
\includegraphics[width=0.33\textwidth]{Figures/grid1119.pdf} 
\itkcaption[ANTS Large Deformation]{The original goal of ANTS was to
  develop public, open source large deformation image registration.
  This is a classic example showing the progress of deforming a half
  C to a full C along a geodesic diffeomorphism.  The deforming 
  grid accompanies each deformed image.}
\vspace{-0.1in}
\label{fig:chalf}
\end{figure}

@@

\subsection{ANTS Scripts} 
The ANTS/Scripts directory contains (hopefully) user-friendly
wrappings of ANTS tools that enable higher-level error checking and
combinations of basic ANTS functions.  These scripts are called as
\verb sh ~ \verb antsscriptname.sh ~ and provide usage when called
from the command line.  For instance, try \verb sh ~ \verb ants.sh .
Additionally, if you run ants in tcsh shells, you will need to call it
as follows:
\begin{verbatim}
~/Code/ANTSSourceForgeSVN/bin/ANTS 2 -m PR\[r16slice.nii,r64slice.nii,1,2\]
\end{verbatim}
instead of
\begin{verbatim}
~/Code/ANTSSourceForgeSVN/bin/ANTS 2 -m PR[r16slice.nii,r64slice.nii,1,2]
\end{verbatim}
because tcsh does no interpret the brackets as intended.  Using the
bash shell script interface also avoids this.  The key change involves using the slash to force tcsh to 
interpet the bracket literally. 
\newpage
\section{ImageRegistration with ANTS}
\begin{wrapfigure}{r}{0.5\textwidth}
\framebox[0.5\textwidth]{
\begin{minipage}[t]{0.45\textwidth}
{\bf Quick Start:} {\color{blue}call ants.sh (from ANTS/Scripts/) to get usage and apply a normalization to some of your existing data.  It is instructional to read the script, modify some of the parameters and re-run to witness the effect of your changes -- image registration is an art as well as science.  Many other ``ready to go'' scripts are available in ANTS/Scripts. The user must set his/her ANTSPATH environment variable  -- which points to the location of ANTS binaries -- within these scripts or in their environment.  Note: All ANTS programs provide usage when called from the command line.  Most require the image dimension to be specified as the first parameter.  E.g. 
  ImageMath ImageDimension 
where ImageDimenion is 2 or 3.}
\end{minipage}}
\end{wrapfigure}
There are two general applications of image registration. The first application is transforming labeled data from a template image space into an individual space. This strategy is important when appearance alone is not enough to locate a structure as, for example, in the case of hippocampus segmentation. The template provides a prediction of the hippocampus-amygdala boundary.
The second application operates in the "inverse" direction of the first: instead of mapping template to individual, we map individual(s) to the template. Voxel-based population studies of either functional or structural variables depend on mapping to a template space. The common coordinate system enables a statistical evaluation of the likelihood of consistent activation across a group or, in other contexts, the differences in anatomy between two groups.

@@

\subsection{Optimal Template Construction with ANTS}
A useful script for setting up optimal template construction
 is available in ANTS/Scripts.  This method is described in \cite{Avants2004,Avants2006d,Kim2008,Yushkevich2009,Avants2009c}, 
where more recent publications are more descriptive and incorporate 
more recent features and optimization strategies. 
Two versions of this algorithm are available : serial and parallel.  
Data and a script for building an average face template is available 
the ANTS/Examples/Data directory.  Two key points about optimal templates:
1. The outcome stabilizes at around ten images, for most populations.
That is, if we randomly select images from individuals within the same 
demographic distribution, we will end up with very similar average images.  
2.  Optimality, for the ANTS-SyN approach, is defined by the minimum shape 
and appearance distance image. 
See the template page for some examples of this and users may 
contact ANTS developers for previously derived template 
images that may be useful.

@@

\newpage

@@

CreateWarpedGridImage 2 PSEtestInverseWarp.nii grid1.nii
CreateWarpedGridImage 2 PSEtestWarp.nii grid2.nii
\end{verbatim}
This example should run on the downloaded ANTS data so you may see the results.
%\input{ANTSLabeledDataExample}
%\input{ANTSMaskedImageRegistrationExample}
\newpage
\section{Application to Studies}
\subsection{Brain Mapping in the Presence of Lesions}
Many cases violate the basic assumption of a one-to-one mapping from a
template image to a target image. Brain lesions caused by stroke or
traumatic brain injury are a common instance of this.

@@

\subsection{Statistics with ANTS and {\bf R}}
{\bf R} is an open-source, cutting-edge statistical package 
used by statisticians world-wide.   ANTS is designed to 
interface with {\bf R} via an ImageMath tool that converts 
image sets into an {\bf R} compatible format.  
Three steps are required.  First, convert image data into 
vector and matrix formats through a mask of your statistical 
region of interest.  Second, read that data into {\bf R} and 
run your statistics.  Third, write out the data from {\bf R} and 
convert back to an image.  Here are the steps, in code examples,
\begin{enumerate}
\item Create a vector for your output: 
\begin{verbatim} 
ImageMath 3 testresultvec.nii ConvertImageSetToMatrix 1 mask.nii image1.nii
\end{verbatim}
Here, we assume all images are called image1.nii, image2.nii, ... , imageN.nii so we 
can use a wildcard to pass to ImageMath and 
create a matrix for all your input data: 
\begin{verbatim} 
 ImageMath 3 datamatrix.nii ConvertImageSetToMatrix 1 mask.nii image*.nii 
\end{verbatim}
\item   Run {\bf R} by reading your data above, performing whatever statistics you want and then 
writing out.  
\item   Convert the test result back to an image.  
\begin{verbatim} 
ImageMath 3 testresult.nii ConvertVectorToImage mask.nii testresultvec.nii
\end{verbatim}
\end{enumerate}
Most statistical requirements may be met with this setup. 
\newpage
\section{Dependencies and Related Software}
\subsection{Dependencies and Compilation}
ANTS depends on the most recent version of the Insight ToolKit (ITK)
-- www.itk.org.   File types that may be read and written are 
all of those available in itk and a few more \href{http://picsl.upenn.edu/ANTS/ioants.php}{http://picsl.upenn.edu/ANTS/ioants.php}.

@@

\subsection{Pipelining with ANTS}
Use PipeDream for automating large-scale processing:\newline
\href{https://sourceforge.net/projects/neuropipedream/}{PipeDream
  Homepage :
  https://sourceforge.net/projects/neuropipedream/} \newline PipeDream
is integrated with ANTS and automates more complex studies such as
multivariate diffusion tensor and T1 cortical thickness studies,
longitudinal mapping and reconstruction of nifti images from Dicom.
\newpage
\section{Annotated Bibliography}
The original statement of the symmetric normalization and template
construction methodology was given in \cite{Avants2004} with a more 
recent template paper here \cite{Avants2009c}. 
A follow up study that used landmark guidance to compare the chimpanzee cortex to
the human cortex was published here \cite{Avants2006} -- this study
used {\em in vivo} MRI and template-based normalization to confirm
volumetric numbers derived from an early 20th century post-mortem
study comparing one human and one chimp.  This conference article has
some additional detail and alternative updates to the methodology, in
particular application to shape-based interpolation
\cite{Avants2005b}.  Network based studies were performed here
\cite{duda08miccai,duda08cvpr}.  The main SyN paper is here
\cite{Avants2008}.  Applications to neurodegeration are here
\cite{Avants2005,Avants2008a,Grossman2008,Avants2009,Das2009,Yushkevich2009,Massimo2009}.
Hippocampus focused work is here \cite{Pluta2009,Yushkevich2009}.  The
main evaluation papers include \cite{Avants2008} and \cite{Klein2009}
for the cortex and deep brain structures whereas \cite{Pluta2009}
evaluates the use of automated and semi-automated normalization for
high-throughput hippocampus morphometry.  An additional evaluation
paper is being developed.
